FHJ: A Formal Model for Resolving Unintended Conflicts in Multiple Inheritance

### 2017-06-15 Aplas17/sections/Overview.tex
temp fix

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {...} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is surely for composing the features from
components, and to achieve code reuse. Multiple inheritance is supported by many mainstream OO
languages. Nevertheless at this point, \lstinline|DrawableDeck| has to throw a compile
error, since the two \lstinline|draw()| methods cause a conflict, though accidentally.

@@

\paragraph{IV. Method exclusion/renaming in traits.} Some trait models
support method exclusion/renaming. Those features
provided can eliminate conflicts, although most
programming languages do not support them. In a traditional OO system,
they more or less break the subtyping. Moreover, in
contrast with exclusion, renaming can indeed preserve both conflicted
behaviours, however, it is cumbersome in practice, as introducing new
names can affect other code blocks. \\

What we really expect from the language is that we keep both methods
without renaming, and the type checker does not complain about 
inheritance. But we need to find an approach to \emph{disambiguate on method
calls statically}.
\bruno{Perhaps show the solution in MIM here (i.e Section 2.4), and then move to problem
2 (which is being discussed next with shuffleAndDraw)}

Certainly the compiler can ignore the conflict when
\lstinline|DrawableDeck| is declared, but once an object of
\lstinline|DrawableDeck| is created, a method call for
\lstinline|draw()| on that object is ambiguous, due to dynamic
dispatch. Nonetheless, we can adopt static dispatch for
disambiguating. Some languages like C++ use qualified names in that
way:

@@

DrawableDeck d = new DrawableDeck();
d.Drawable::draw();       // calling draw() in Drawable
((Drawable) d).draw(); // calling draw() in Drawable
func(d);                 // calling draw() in Drawable
\end{lstlisting}\vspace{3pt}
Thus we have: \paragraph{V. Static dispatch.} Static dispatch finds out and invokes the most specific method ``by need''.

On the other hand, we also need dynamic dispatch as it is essential and widely used in object-oriented programming.
C++ has the flexibility for choosing either way of dispatch by the ``virtual'' keyword.
Unfortunately, this approach is still unsatisfactory regarding code reuse. For instance, here we redefine \lstinline|Deck| to support
both \lstinline|draw()| and another operation called \lstinline|shuffleAndDraw()|:
\vspace{3pt}\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffleAndDraw() {
    shuffle();
    draw();
  }
  ...
}
\end{lstlisting}\vspace{3pt}
\lstinline|shuffleAndDraw()| is a representative method that invokes \lstinline|draw()| in its definition. In principle, we want
that invocation to use dynamic dispatch, because a programmer may define a subtype of \lstinline|Deck|, and override \lstinline|draw()|:
\vspace{3pt}\begin{lstlisting}
interface LoggingDeck {
  void draw() { // overriding
    Stack<Card> cards = this.getStack();
    if (!cards.isEmpty()) {
      Card card = cards.pop();
      println("The card is: " + card.toString());
      ...
    } else {
      println("Empty deck.");
    }
  }
}
\end{lstlisting}\vspace{3pt}
Usually we want to reuse the code of \lstinline|shuffleAndDraw()| during inheritance, hence dynamic dispatch is necessary, otherwise
programmers have to override all the other methods that invoke \lstinline|draw()|. However, as seen before, dynamic dispatch can cause
ambiguity if we have:
\vspace{3pt}\begin{lstlisting}
interface DrawableLoggingDeck extends Drawable, LoggingDeck {...}

@@

\subsection{Solution in \MIM: \dispatchnamecaptical}\label{subsec:solutionmim}
The \MIM{} model uses \dispatchnameit{} for method lookup. A qualified
method invocation, for instance, \lstinline|e.I::m()|, is read as
``finding the most specific method \lstinline|m| along the path
\lstinline|I|''. The meaning of ``along the path \lstinline|I|'' is
that, if the result of \dispatch{} is \lstinline|J.m()| for some \lstinline|J|, then such a \lstinline|J| must be a super type of \lstinline|e|'s dynamic type, and \lstinline|J| has a subtyping relation with \lstinline|I| (either \lstinline|J <: I| or \lstinline|J >: I|). Intuitively, the most specific \lstinline|m()| must be from branch \lstinline|I|, but it can be an overridden version after \lstinline|I| like dynamic dispatch. The formal definition will be introduced later. On the other hand, \lstinline|e.m()| where \lstinline|e| has static type \lstinline|I|, simply behaves the same as \lstinline|e.I::m()| in our model. It is just like the difference between implicit and explicit casts.

Such a dispatch make uses of both the static type and the dynamic type of the receiver. Intuitively, the static type specifies one branch of the method to avoid ambiguity, and the dynamic type finds the latest version on that branch. It may still introduce ambiguity when there are multiple paths from the static type to the dynamic type, and those paths cause conflicts. To prevent this, we disallow this kind of conflict to ensure unambiguity. That is to say, we do not allow two methods to override a same base method when they cause conflicts. This is natural, as they are just two versions of the same operation, hence it is no longer an ``unintentional'' conflict in that way.

With the old example, below code meets our expectation:
\vspace{3pt}\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffleAndDraw() {
    this.Deck::shuffle();
    this.Deck::draw();
  }
  ...
}
\end{lstlisting}\vspace{3pt}
It guarantees that \lstinline|shuffleAndDraw()| are calling \lstinline|shuffle()| and \lstinline|draw()| from its own branch, so that the namesakes
from other branches will not cause conflicts. Now \lstinline|d.shuffleAndDraw()| is no longer ambiguous.

>>>>

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {...} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is surely for composing the features from various 
components, and to achieve code reuse. Multiple inheritance is supported by many mainstream OO
languages. Nevertheless at this point, \lstinline|DrawableDeck| has to throw a compile
error, since the two \lstinline|draw()| methods cause a conflict, though accidentally.

@@

\paragraph{IV. Method exclusion/renaming in traits.} Some trait models
support method exclusion/renaming. Those features
provided can eliminate conflicts, although most
programming languages do not support them. In a traditional OO system,
they more or less break the subtyping. Moreover, in
contrast with exclusion, renaming can indeed preserve both conflicted
behaviours, however, it is cumbersome in practice, as introducing new
names can affect other code blocks.

\paragraph{V. Static dispatch.}
What we really expect from the language is keeping both methods
without renaming, and the type checker does not complain. 
But we need to find an approach to \emph{disambiguate on method
calls statically}.
\bruno{Perhaps show the solution in MIM here (i.e Section 2.4), and then move to problem
2 (which is being discussed next with shuffleAndDraw)}

Certainly the compiler can ignore the conflict when
\lstinline|DrawableDeck| is declared, but once an object of
\lstinline|DrawableDeck| is created, a method call for
\lstinline|draw()| on that object would be ambiguous. Nonetheless, we can adopt static dispatch for
disambiguating. Some languages like C++ use qualified names in that
way:

@@

DrawableDeck d = new DrawableDeck();
d.Drawable::draw();       // calling draw() in Drawable
((Drawable) d).draw(); // calling draw() in Drawable
func(d);                 // calling draw() in Drawable
\end{lstlisting}\vspace{3pt}
Static dispatch in C++ finds out and invokes the most specific method ``by need''.

\subsection{Problem 2: Dynamic Dispatch for Code Reuse}
On the other hand, we also need dynamic dispatch as it is essential and widely used in object-oriented programming.
C++ has the flexibility for choosing either way of dispatch by the \kwvirtual{} keyword.
Unfortunately, this approach is still unsatisfactory regarding code reuse. For instance, here we redefine \lstinline|Deck| to support
both \lstinline|draw()| and another operation called \lstinline|shuffleAndDraw()|:
\vspace{3pt}\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffleAndDraw() {
    shuffle();
    draw();
  }
  ...
}
\end{lstlisting}\vspace{3pt}
\lstinline|shuffleAndDraw()| is a representative method that invokes \lstinline|draw()| in its definition. In principle, we want
that invocation to use dynamic dispatch, because a programmer may define a subtype of \lstinline|Deck|, and override \lstinline|draw()|:
\vspace{3pt}\begin{lstlisting}
interface LoggingDeck {
  void draw() { // overriding
    Stack<Card> cards = this.getStack();
    if (!cards.isEmpty()) {
      Card card = cards.pop();
      println("The card is: " + card.toString());
      ...
    } else {
      println("Empty deck.");
    }
  }
}
\end{lstlisting}\vspace{3pt}
Usually we want to reuse the code of \lstinline|shuffleAndDraw()| during inheritance, hence dynamic dispatch is necessary, otherwise
programmers have to override all the other methods that invoke \lstinline|draw()|. However, as seen before, dynamic dispatch can cause
ambiguity if we have:
\vspace{3pt}\begin{lstlisting}
interface DrawableLoggingDeck extends Drawable, LoggingDeck {...}

@@

\subsection{Solution in \MIM: \dispatchnamecaptical}\label{subsec:solutionmim}
The \MIM{} model uses \dispatchnameit{} for method lookup. A qualified
method invocation, for instance, \lstinline|e.I::m()|, is read as
``finding the most specific method \lstinline|m| along the path
\lstinline|I|''. The meaning of ``along the path \lstinline|I|'' is
that, if the result of \dispatch{} is \lstinline|J.m()| for some \lstinline|J|, then such a \lstinline|J| must be a super type of \lstinline|e|'s dynamic type, and \lstinline|J| has a subtyping relation with \lstinline|I| (either \lstinline|J <: I| or \lstinline|J >: I|). Intuitively, the most specific \lstinline|m()| must be from branch \lstinline|I|, but it can be an overridden version after \lstinline|I| because of dynamic dispatch. The formal definition will be introduced later. On the other hand, \lstinline|e.m()| where \lstinline|e| has static type \lstinline|I|, simply behaves the same as \lstinline|e.I::m()| in our model. It is just like the difference between implicit and explicit casts.

Such a dispatch algorithm make uses of both the static type and the dynamic type of the receiver. Intuitively, the static type specifies one branch of the method to avoid ambiguity, and the dynamic type finds the latest version on that branch. It may still introduce ambiguity when there are multiple paths from the static type to the dynamic type, and those paths cause conflicts. To prevent this, we disallow this kind of conflict to ensure unambiguity. That is to say, we do not allow two methods to override a same base method when they cause conflicts. This is natural, as they are just two versions of the same operation, hence it is no longer an ``unintentional'' conflict in that way.

With the old example, the code below meets our expectation:
\vspace{3pt}\begin{lstlisting}
interface Deck {
  void draw() {...}
  void shuffleAndDraw() {
    this.Deck::shuffle();
    this.Deck::draw();
  }
  ...
}
\end{lstlisting}\vspace{3pt}
It guarantees that \lstinline|shuffleAndDraw()| calls \lstinline|shuffle()| and \lstinline|draw()| from its own branch, so that the namesakes
from other branches will not cause conflicts. Now \lstinline|d.shuffleAndDraw()| is no longer ambiguous.
